---
title: "Organelle_analysis"
author: "Fiona M. Love"
date: "2025-04-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(readbitmap)
library(RImageJROI)
library(spatstat.geom)
library(sf)
library(dplyr)
library(terra)
library(ggplot2)
library(ggpubr)
library(patchwork)

```

#ANALYSIS

Select ImageJ Macro output directory
*R VERSION BUG ON WINDOWS 10* - WON'T OPEN SELECTION WINDOW UNTIL AFTER file.choose() HAS BEEN RUN
https://stackoverflow.com/questions/77984544/choose-dir-function-does-not-work-in-r-v4-3-2
```{r data_location}
dir <- choose.dir() #WINDOWS ONLY - TODO Mac support with tk option - https://stackoverflow.com/questions/48218491/os-independent-way-to-select-directory-interactively-in-r
```

Identify conditions for matching and iterating over ROI/measurement files
```{r conditions}
files <- list.files(dir, pattern = ('\\.csv$|\\.zip$'), recursive = FALSE, include.dirs = FALSE)
conditions <- unique(gsub('_analysis.csv||_ROIs.zip', '', files))
```

Identify features segmented
```{r features}
has_cells <- TRUE #REQUIRED
has_nuclei <- TRUE
has_Golgi <- TRUE
has_peroxisomes <- FALSE
has_ER <- FALSE
has_mito <- FALSE
has_lyso <- FALSE
has_bacteria <- TRUE

has_pairwise_mcs <- FALSE
has_multi_mcs <- FALSE
```

Identify files for each condition
```{r file_matching}
#currently reading in all data - consider moving into loop if too much in memory at once
conditions.data <- lapply(conditions, function(condition){
  print(condition)
  c <- list()
  measurements <- read.csv(paste0(dir, '\\', condition, '_analysis.csv'))
  measurements$ROI_name <- unlist(regmatches(measurements$ID, regexec("(?<=:).+(?=:)?", measurements$ID, perl = T)))
  #TODO - this might break if trailing : missing - dependent on single/multi channel image - should be consistent with macro
  measurements.cells = measurements[measurements$Type == 'cell',]
  measurements.other = measurements[measurements$Type != 'cell',]
  
  ROIs = read.ijzip(paste0(dir, '\\', condition, '_ROIs.zip'), names = TRUE)
  ROIs.cells <- ROIs[measurements[measurements$Type == 'cell',]$ROI_name]
  ROIs.other <- ROIs[measurements[measurements$Type != 'cell',]$ROI_name]
  remove(ROIs)
  remove(measurements)
  
  c[['measurements.cells']] <- measurements.cells
  c[['measurements.other']] <- measurements.other
  c[['ROIs.cells']] <- ROIs.cells
  c[['ROIs.other']] <- ROIs.other
  return(c)
})
names(conditions.data) = conditions
```

Defining function for converting ImageJ ROIs into sf polygons
```{r st_polygon_from_coords}
  st_polygon_from_coords <- function(ijroi){
    coords <- ijroi$coords
    #loop
    coords <- coords[c(1:nrow(coords), 1),]
    coords.list <- list(coords)
    tryCatch(expr = st_polygon(coords.list), finally = st_polygon(coords.list[rev(1:length(coords.list))]))
  }
```

BATCH LOOP
```{r batch loop}
per_cell.all = do.call(rbind, lapply(conditions, function(condition){
  
  c = conditions.data[[condition]]
  measurements.cells = c[['measurements.cells']]
  measurements.other = c[['measurements.other']]
  ROIs.cells = c[['ROIs.cells']]
  ROIs.other = c[['ROIs.other']]
  
  sf.cells = lapply(ROIs.cells, st_polygon_from_coords)
  sf.other = lapply(ROIs.other, st_polygon_from_coords)
  
  sf.per_cell = lapply(sf.cells, function(c){ 
    lapply(sf.other, function(o){
      #st_within(o, c, sparse = FALSE)
      st_intersects(o, c, sparse = FALSE)#more lenient for structures that might overlap with segmented cell boundary
    })
  })
  sf.per_cell.filtered = lapply(1:length(sf.per_cell), function(c){ which(sf.per_cell[[c]] == TRUE) })
  #assigning names later
  
  
  
  #PER-CELL ANALYSES
  
  names(sf.per_cell.filtered) = names(sf.per_cell)
  other_to_cells = data.frame(other = c(), cell = c())
  for (c in names(sf.cells)){ 
    for (o in names(sf.other)){
      if(o %in% names(sf.per_cell.filtered[[c]])){
        other_to_cells = rbind(other_to_cells, data.frame(other = o, cell = c))
      }
    }
  }
  measurements.other$cell = sapply(measurements.other$ROI_name, function(n){ other_to_cells[other_to_cells$other == n,]$cell })
  
  #handling ROIs without a cell, and ROIs assigned to multiple cells by overlap
  measurements.other = do.call(rbind, lapply(1:nrow(measurements.other), function(i){
    row = measurements.other[i,]
    celllist = row[,"cell"]
    if(length(celllist[[1]]) == 0){ row$cell = NA }#catch lists with empty string as first item
    if(length(celllist[[1]]) == 1){ row$cell = celllist[[1]] }
    else{
      #duplicate row for each cell in list
      row = do.call(rbind, lapply(celllist[[1]], function(c){
        row$cell = c
        row
      }))
    }
    return(row)
  }))
  
  other_to_cells$type = sapply(other_to_cells$other, function(o){ measurements.other[measurements.other$ROI_name == o,]$Type[1]})#should be unique; adding [1] for output formatting 
  
  #identify infected cells
  cells.contain = data.frame(cell = unique(other_to_cells$cell))
  cells.contain$nucleus = sapply(cells.contain$cell, function(c){ "nucleus" %in% other_to_cells[other_to_cells$cell == c, ]$type })
  cells.contain$golgi = sapply(cells.contain$cell, function(c){ "golgi" %in% other_to_cells[other_to_cells$cell == c, ]$type })
  cells.contain$bac = sapply(cells.contain$cell, function(c){ "bac" %in% other_to_cells[other_to_cells$cell == c, ]$type })
  
  infected = (cells.contain %>% filter(nucleus == TRUE, golgi == TRUE, bac == TRUE))$cell
  uninfected = (cells.contain %>% filter(nucleus == TRUE, golgi == TRUE, bac == FALSE))$cell
  valid = c(infected, uninfected)
  
  measurements.other$infected = measurements.other$cell %in% infected
  measurements.other$valid = measurements.other$cell %in% c(infected, uninfected)
  
  #MDR analysis
  #TODO - probably faster to read in cell binaries rather than rasterize, but would need to reassign to ROIs
  cells.rasters = lapply(valid, function(c){ 
    r = extend(rasterize(vect(sf.cells[[c]]), rast(vect(sf.cells[[c]]),res=1)), 1)
    crs(r) <- 'local'
    return(r)
  })
  names(cells.rasters) = valid
  cells.pixels = lapply(valid, function(c){ as.data.frame(cells.rasters[[c]], xy = TRUE) })
  names(cells.pixels) = valid
  
  cells.nuc_cent = do.call(rbind, lapply(valid, function(c){ 
    cn = which(names(sf.other) %in% other_to_cells[other_to_cells$cell == c & other_to_cells$type == 'nucleus',]$other)
    if(length(cn) > 1){
      nc = st_centroid(st_geometrycollection(sf.other[cn]))
      return(data.frame(Cent_X_pix = nc[1], Cent_Y_pix = nc[2], cell = c))
    }
    else{
      return(measurements.other[measurements.other$cell == c & measurements.other$Type == 'nucleus', c('Cent_X_pix', 'Cent_Y_pix', 'cell')])
    }
  }))
  
  cells.golgi_cent = do.call(rbind, lapply(valid, function(c){ 
    cg = which(names(sf.other) %in% other_to_cells[other_to_cells$cell == c & other_to_cells$type == 'golgi',]$other)
    if(length(cg) > 1){
      gc = st_centroid(st_geometrycollection(sf.other[cg]))
      return(data.frame(Cent_X_pix = gc[1], Cent_Y_pix = gc[2], cell = c))
    }
    else{
      return(measurements.other[measurements.other$cell == c & measurements.other$Type == 'golgi', c('Cent_X_pix', 'Cent_Y_pix', 'cell')])
    }
  }))
  
  
  #for Pixel distance Golgi-based MDR
  
  organelles.spatvectors = lapply(names(sf.other), function(o){ 
    v = vect(sf.other[[o]]) 
    crs(v) <- 'local'
    return(v)
  })
  names(organelles.spatvectors) = names(sf.other)
  for(org in names(organelles.spatvectors)){
    organelles.spatvectors[[org]]$rast_val = 2
  }#TODO - organelle type metadata?
  
  #TODO URGENT (for Tom) - iterate for every organelle type, then rbind for organelles.rasters below rather than regenerating
  golgi.rasters = lapply(names(cells.pixels), function(cell){
    cell.golgi = other_to_cells[other_to_cells$cell == cell & other_to_cells$type == 'golgi', "other"]
    golgi.spatvector = vect(organelles.spatvectors[cell.golgi])
    g = extend(rasterize(golgi.spatvector, rast(golgi.spatvector, res = 1, crs = 'local'), field = 'rast_val'), 1)#extend might not be necessary here - check
    #ERROR HERE - TODO 8/4
    #crs(g) <- 'local'
    g = extend(g, cells.rasters[[cell]]) #changing extent of raster to fit corresponding cell for distance calculations
    return(g)
  })
  names(golgi.rasters) = names(cells.pixels)
  #also ensure cell raster extents encompass entire golgi assigned to that cell:
  cells.rasters = lapply(names(cells.rasters), function(cell){
    c = cells.rasters[[cell]]
    if(cell %in% names(golgi.rasters)){#shouldn't be necessary...?
      c = extend(c, golgi.rasters[[cell]])
    }
    return(c)
  })
  names(cells.rasters) = valid#names lost above
  
  
  for(name in names(cells.pixels)){
    cell = cells.pixels[[name]]
    #distance to nucleus/PM and standard MDR
    cell$dist_to_nuc = sapply(1:nrow(cell), function(px){ 
      px_x = cell[px, 'x']
      px_y = cell[px, 'y']
      cent = cells.nuc_cent[cells.nuc_cent$cell == name,]
      return(abs(sqrt((px_x - cent$Cent_X_pix[1])^2 + (px_y - cent$Cent_Y_pix[1])^2)))
      # #index for formatting - multiple golgi/nuclei handled above in centroid calculation
    })
    cell$dist_to_PM = as.data.frame(mask(distance(cells.rasters[[name]], target = 1), cells.rasters[[name]]))[,1]#distance to nearest PM pixel (outside cell)
    cell$norm = cell$dist_to_nuc/(cell$dist_to_nuc + cell$dist_to_PM)
    
    cell$angle = sapply(1:nrow(cell), function(px){ 
      px_x = cell[px, 'x']
      px_y = cell[px, 'y']
      cent = cells.nuc_cent[cells.nuc_cent$cell == name,]
      return((atan2((px_y - cent$Cent_Y_pix[1]), (px_x - cent$Cent_X_pix[1]))*180)/pi)#index for formatting - multiple golgi/nuclei handled above in centroid calculation
      #TODO - combine with dist_to_nuc
    })
    
    #Golgi-based MDR
    cell$dist_to_golgi = sapply(1:nrow(cell), function(px){ 
      px_x = cell[px, 'x']
      px_y = cell[px, 'y']
      cent = cells.golgi_cent[cells.golgi_cent$cell == name,]
      return(abs(sqrt((px_x - cent$Cent_X_pix[1])^2 + (px_y - cent$Cent_Y_pix[1])^2)))#index for formatting - multiple golgi/nuclei handled above in centroid calculation
    })
    cell$norm_to_golgi = cell$dist_to_golgi/(cell$dist_to_golgi + cell$dist_to_PM)
    
    #Golgi-based MDR NEAREST PIXEL
    cell$dist_to_nearest_golgi = as.data.frame(mask(distance(golgi.rasters[[name]], target = NA), cells.rasters[[name]]))[,1]#distance to nearest golgi pixel
    cell$norm_to_nearest_golgi = cell$dist_to_nearest_golgi/(cell$dist_to_nearest_golgi + cell$dist_to_PM)

    cells.pixels[[name]] = cell
  }
  

  
  organelles.rasters = lapply(names(sf.other), function(o){ 
    r = extend(rasterize(organelles.spatvectors[[o]], rast(organelles.spatvectors[[o]],res=1), field = 'rast_val'), 1) 
    crs(r) <- 'local'
    return(r)
  })
  names(organelles.rasters) = names(sf.other)
  organelles.pixels = lapply(names(sf.other), function(o){ as.data.frame(organelles.rasters[[o]], xy = TRUE) })
  names(organelles.pixels) = names(sf.other)
  
  for(n in names(organelles.pixels)){
    organelles.pixels[[n]]$name = n
    organelles.pixels[[n]]$type = measurements.other[measurements.other$ROI_name == n,]$Type[1]
  }
  
  for(name in names(cells.pixels)){
    cell = cells.pixels[[name]]
    cell$golgi = FALSE
    cell$nucleus = FALSE
    cell$bac = FALSE
    cell$ng = FALSE
    cell$gb = FALSE
    cell$bn = FALSE
    cell.org = organelles.pixels[which(names(organelles.pixels) %in% other_to_cells[other_to_cells$cell == name,]$other)]
    for (o in cell.org){
      for(i in 1:nrow(o)){
        o.px = o[i,]
        cell.px = which(cell$x == o.px$x & cell$y == o.px$y)
        cell[cell.px, o.px$type[1]] = TRUE
      }
    }
    cell$name = name
    cell$infected = name %in% infected
    cells.pixels[[name]] = cell
  }
  
  
  cells.pixels.pooled = do.call(rbind, cells.pixels)
  
  per_cell = do.call(rbind, lapply(names(cells.pixels), function(n){
    cell.condition = condition
    cell.px = cells.pixels.pooled[cells.pixels.pooled$name == n,]
    cell.golgi_norm = mean(cell.px[cell.px$golgi == TRUE,]$norm)
    cell.bac_norm = mean(cell.px[cell.px$bac == TRUE,]$norm)
    cell.bac_norm_to_golgi = mean(cell.px[cell.px$bac == TRUE,]$norm_to_golgi)
    cell.bac_dist_to_nearest_golgi = mean(cell.px[cell.px$bac == TRUE,]$dist_to_nearest_golgi)
    cell.bac_norm_to_nearest_golgi = mean(cell.px[cell.px$bac == TRUE,]$norm_to_nearest_golgi)
    cell.golgi_area = nrow(cell.px[cell.px$golgi == TRUE,])#TODO - check against ImageJ measurements
    cell.gb_area = nrow(cell.px[cell.px$gb == TRUE,])
    
    cell.golgi_count = nrow(other_to_cells[other_to_cells$cell == n & other_to_cells$type == 'golgi',])
    cell.nucleus_count = nrow(other_to_cells[other_to_cells$cell == n & other_to_cells$type == 'nucleus',])
    
    cell.name = cell.px$name[1]
    cell.infected = cell.px$infected[1]
    
    cell.area = measurements.cells[measurements.cells$ROI_name == cell.name,]$Area[1]
    cell.circ = measurements.cells[measurements.cells$ROI_name == cell.name,]$Circ[1]
    
    cell.df = data.frame(
      cond = cell.condition,
      name = cell.name,
      infected = cell.infected,
      golgi_norm = cell.golgi_norm,
      bac_norm = cell.bac_norm,
      bac_norm_to_golgi = cell.bac_norm_to_golgi,
      bac_dist_to_nearest_golgi = cell.bac_dist_to_nearest_golgi,
      bac_norm_to_nearest_golgi = cell.bac_norm_to_nearest_golgi,
      golgi_area = cell.golgi_area,
      gb_area = cell.gb_area,
      cell_area = cell.area,
      cell_circ = cell.circ,
      golgi_count = cell.golgi_count,
      nucleus_count = cell.nucleus_count
    )
  }))
  return(per_cell)
  remove(per_cell)
}))
```

Post-processing variables
```{r post_cols}
per_cell.all$date = sapply(per_cell.all$cond, function(s){strsplit(s, '_')[[1]][1]})
per_cell.all$cell_line = as.factor(sapply(per_cell.all$cond, function(s){strsplit(s, '_')[[1]][2]}))
per_cell.all$strain = as.factor(sapply(per_cell.all$cond, function(s){strsplit(s, '_')[[1]][3]}))
per_cell.all$number = sapply(per_cell.all$cond, function(s){strsplit(s, '_')[[1]][4]})
```

Write output
```{r save_CSV}
write.csv(per_cell.all, paste0(dir, '\\organelle_analysis\\analysis_output.csv'))
```


#PLOTS
```{r plots}
ggplot(per_cell.all[per_cell.all$infected == T,], aes(x = strain, y = bac_norm_to_nearest_golgi, group = strain)) +
  geom_dotplot(binwidth = 0.01, binaxis='y', stackdir='center') + geom_boxplot(fill = NA, outliers = F) + facet_wrap(~cell_line) + stat_compare_means()

ggplot(per_cell.all[per_cell.all$infected == T,], aes(x = strain, y = bac_dist_to_nearest_golgi, group = strain)) +
  geom_dotplot(binwidth = 0.01, binaxis='y', stackdir='center') + geom_boxplot(fill = NA, outliers = F) + facet_wrap(~cell_line) + stat_compare_means()



ggplot(per_cell.all[per_cell.all$infected == T & per_cell.all$strain == 'CGS001',], aes(x = cell_line, y = bac_norm_to_golgi, group = cell_line)) +
  geom_dotplot(binwidth = 0.01, binaxis='y', stackdir='center') + geom_boxplot(fill = NA, outliers = F) + stat_compare_means()

ggplot(per_cell.all[per_cell.all$infected == T,], aes(x = strain, y = bac_norm, group = strain)) +
  geom_dotplot(binwidth = 0.01, binaxis='y', stackdir='center') + geom_boxplot(fill = NA, outliers = F) + facet_wrap(~cell_line) + stat_compare_means()

ggplot(per_cell.all[per_cell.all$infected == T,], aes(x = strain, y = golgi_norm, group = strain)) +
  geom_dotplot(binwidth = 0.01, binaxis='y', stackdir='center') + geom_boxplot(fill = NA, outliers = F) + facet_wrap(~cell_line) + stat_compare_means()
```
