---
title: "Organelle Analysis - ImageJ Metrics"
author: "Fiona M. Love"
date: "2025-05-09"
output: html_document
---

#USER PARAMETERS

*file name parsing* - each section of the file name delineated by underscores (_) can be passed as a variable representing e.g. timepoint, cell line, experimental condition, etc.  Indicate which place each one is in here, ensuring there are no spaces or other special characters (besides underscores) in the name.
e.g. for something like "HeLa_Ctrl_cell1":
up['cell_line'] <- 1
up['condition'] <- 2
up['cell'] <- 3

*normalise_organelles_to* - Which measurement to use to normalise organelle areas.  Options are 'cell', 'cytoplasm' (cell area - nucleus area), 'both', or 'none'. 'Cytoplasm' or 'both' will only work if nuclei were measured in the ImageJ analysis step, and nucleus will still be normalised to cell area unless 'none' is selected.  Organelle contacts are normalised to the cumulative area of the organelle types involved.



```{r user_params}
up <- list()#leave this line as-is - this variable will be used to retrieve the values below
up['cell_line'] <- 2
up['clone'] <- 3
up['weights'] <- 4
up['timepoint'] <- 5
up['image'] <- 6
up['cell_number'] <- 10

normalise_organelles_to <- 'cell'

```









```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(ggplot2)
library(ggpubr)
library(patchwork)
library(corrr)
library(ggcorrplot)
library(factoextra)

```

#ANALYSIS

*Select ImageJ Macro output directory*
_R VERSION BUG ON WINDOWS 10_ - WON'T OPEN SELECTION WINDOW UNTIL AFTER file.choose() HAS BEEN RUN
https://stackoverflow.com/questions/77984544/choose-dir-function-does-not-work-in-r-v4-3-2
```{r data_location}
dir <- choose.dir() #WINDOWS ONLY - TODO Mac support with tk option - https://stackoverflow.com/questions/48218491/os-independent-way-to-select-directory-interactively-in-r
```

*Identify conditions for matching and iterating over ROI/measurement files*
```{r conditions}
files <- list.files(dir, pattern = ('\\.csv$|\\.zip$'), recursive = FALSE, include.dirs = FALSE)
files <- files[-which(grepl('\\_ALL\\_CELL\\_ROIs\\.zip$|\\_ALL\\_ORGANELLE\\_ROIs\\.zip$', files))]
condcells <- unique(gsub('_results.csv||_POOLED_results.csv||_POOLED_ROIs.zip||_ROIs.zip||_summary_results.csv||_slice_labels.csv', '', files))#updated for cell-by-cell processing
```

*Identify files for each condition/cell*
Can't read in complex ROIs with RImageJROI package - measurements only for now
```{r file_matching}

setClass("Measurements", slots = c(data = "data.frame", cells = "data.frame", other = "data.frame"))

#measurements cleaning/categories
process_measurements <- function(dir, condition, file_pattern){
  df <- read.csv(paste0(dir, '\\', condition, file_pattern))
  df$ROI_name <- unlist(regmatches(df$Label, regexec("(?<=:).+(?=:)?", df$Label, perl = T)))
  #TODO - this might break if trailing : missing - dependent on single/multi channel image - should be consistent with macro
  df.cells = df[df$Type == 'cells',]
  df.other = df[df$Type != 'cells',]
  mes <- new("Measurements", data = df, cells = df.cells, other = df.other)
  return(mes)
}

pooled_measurements <- lapply(condcells, function(c){process_measurements(dir, c, "_POOLED_results.csv")})
names(pooled_measurements) <- condcells
individual_measurements <- lapply(condcells, function(c){process_measurements(dir, c, "_results.csv")})
names(individual_measurements) <- condcells
summary_measurements <- lapply(condcells, function(c){process_measurements(dir, c, "_summary_results.csv")})
names(summary_measurements) <- condcells

```

*Organelle set*
```{r organelles}
#TODO - generate from slice names?
organelles <- unique(unlist(lapply(condcells, function(c){ pooled_measurements[[c]]@other$Type })))
pairwise <- list("ng" = c("nuclei", "Golgi"),#is this used anywhere? #TODO - change ASAP to accommodate higher order contacts
"np" = c("nuclei", "peroxisomes"),
"ne" = c("nuclei", "ER"),
"nm" = c("nuclei", "mitochondria"),
"nl" = c("nuclei", "lysosomes"),
"no" = c("nuclei", "other"),
"gp" = c("Golgi", "peroxisomes"),
"ge" = c("Golgi", "ER"),
"gm" = c("Golgi", "mitochondria"),
"gl" = c("Golgi", "lysosomes"),
"go" = c("Golgi", "other"),
"pe" = c("peroxisomes", "ER"),
"pm" = c("peroxisomes", "mitochondria"),
"pl" = c("peroxisomes", "lysosomes"),
"po" = c("peroxisomes", "other"),
"em" = c("ER", "mitochondria"),
"el" = c("ER", "lysosomes"),
"eo" = c("ER", "other"),
"ml" = c("mitochondria", "lysosomes"),
"mo" = c("mitochondria", "other"),
"lo" = c("lysosomes", "other"))
```


*Combine into table of defined measurements*
```{r data_frame}
metrics <- data.frame(condcell = condcells)

#---custom variables from file name---
for(param in names(up)){
  metrics[param] <- sapply(condcells, function(c){strsplit(c, "_")[[1]][up[[param]]]})
}
#------------

metrics$cell_area <- sapply(condcells, function(c){ pooled_measurements[[c]]@cells$Area })
metrics$cell_circ <- sapply(condcells, function(c){ pooled_measurements[[c]]@cells$Circ. })

for(org in c(organelles, names(pairwise))){#TODO ASAP - higher order
  metrics[paste0(org, "_area")] <- sapply(condcells, function(c){ other = pooled_measurements[[c]]@other;
                                                          other[other$Type == org,]$Area[1] })#should only have one of each - MIGHT BREAK IF PAIRWISE MISSING - CHECK
  metrics[paste0(org, "_area_norm_cell")] <- metrics[paste0(org, "_area")]/metrics$cell_area
  
  metrics[paste0(org, "_count")] <- sapply(condcells, function(c){ other = individual_measurements[[c]]@other;#FROM INDIVIDUALS DF
                                                          nrow(other[other$Type == org,]) })
  
  metrics[paste0(org, "_size_avg")] <- sapply(condcells, function(c){ other = individual_measurements[[c]]@other;#FROM INDIVIDUALS DF
                                                        mean(other[other$Type == org,]$Area) })
  metrics[paste0(org, "_size_sd")] <- sapply(condcells, function(c){ other = individual_measurements[[c]]@other;#FROM INDIVIDUALS DF
                                                          sd(other[other$Type == org,]$Area) })
  
  metrics[paste0(org, "_circ_avg")] <- sapply(condcells, function(c){ other = individual_measurements[[c]]@other;#FROM INDIVIDUALS DF
                                                        mean(other[other$Type == org,]$Circ.) })
  metrics[paste0(org, "_circ_sd")] <- sapply(condcells, function(c){ other = individual_measurements[[c]]@other;#FROM INDIVIDUALS DF
                                                          sd(other[other$Type == org,]$Circ.) })
}

metrics$total_organelle_area <- sapply(condcells, function(c){df <- summary_measurements[[c]]@data; df[df$Slice == 'organelles','Total.Area'][1]})
metrics$total_organelle_area_norm_cell <- metrics$total_organelle_area/metrics$cell_area
metrics$total_contact_area <- sapply(condcells, function(c){df <- summary_measurements[[c]]@data; df[df$Slice == 'pairwise_contact_sites','Total.Area'][1]})
metrics$total_contact_area_norm_cell <- metrics$total_contact_area/metrics$cell_area
metrics$total_contact_area_norm_organelles <- metrics$total_contact_area/metrics$total_organelle_area

if('nuclei_area' %in% colnames(metrics)){
  metrics$cytoplasmic_area <- metrics$cell_area - metrics$nuclei_area
  for(org in c(organelles)){
    metrics[paste0(org, "_area_norm_cyto")] <- metrics[paste0(org, "_area")]/metrics$cytoplasmic_area#TODO - conditional on user params
    #TODO - norm nuclei to cell area instead of cyto
  }
}




```



## Data cleaning ##
```{r}
metrics[is.na(metrics)] <- 0
# #Tom specific: replaces all bacterial measurements with 0 in uninfected conditions
# metrics[which(metrics$infected == 'UNinfected'),] <- metrics[which(metrics$infected == 'UNinfected'),] %>% mutate(across(contains('bacteria'), ~replace(.,1:length(.),0)))
# metrics[which(metrics$infected == '1dpi'),]$infected <- 'infected'
```


## Extra Metrics ##
Optional - passing through additional measurements from ImageJ, with details from file names
```{r}
# pooled.df <- do.call(rbind, lapply(1:length(pooled_measurements), function(p){ df <- pooled_measurements[[p]]@data; df$filename <- names(pooled_measurements)[p]; return(df) }))
# extra_measurements.pooled <- left_join(metrics[, c('infection_time', 'infected', 'bac_line', 'cell_line', 'image', 'cell_number', 'condcell')], pooled.df, join_by(condcell == filename))
# 
# individual.df <- do.call(rbind, lapply(1:length(individual_measurements), function(i){ df <- individual_measurements[[i]]@data; df$filename <- names(individual_measurements)[i]; return(df) }))
# extra_measurements.individual <- left_join(metrics[, c('infection_time', 'infected', 'bac_line', 'cell_line', 'image', 'cell_number', 'condcell')], individual.df, join_by(condcell == filename))
```



