---
title: "Organelle Analysis - all metrics"
author: "Fiona M. Love"
date: "2025-11-11"
output: html_document
---
# ORGANELLE ANALYSIS

## User parameters

*file name parsing* - each section of the file name delineated by underscores (_) can be passed as a variable representing e.g. timepoint, cell line, experimental condition, etc.  Indicate which place each one is in here, ensuring there are no spaces or other special characters (besides underscores) in the name.
e.g. for something like "HeLa_Ctrl_cell1":
up['cell_line'] <- 1
up['condition'] <- 2
up['cell'] <- 3

*binangle* - The angle to use to split the cell into slices for calculating coefficient of asymmetry.  Must divide 360 into an even number of groups - e.g. 5, 6, or 10 is fine, but 7 or 8 won't work.


*use_normalised_metrics* - Whether to use normalised or raw area metrics; TRUE or FALSE (without quotes).  Defaults to TRUE.

*normalise_organelles_to* - Which measurement to use to normalise organelle areas.  Options are 'cell', 'cytoplasm' (cell area - nucleus area), 'both', or 'none'. 'Cytoplasm' or 'both' will only work if nuclei were measured in the ImageJ analysis step, and nuclei will still be normalised to cell area unless 'none' is selected.  Organelle contacts are always normalised to the cumulative area of the organelle types involved.

*distance_metric* - Which method to use to determine how far each organelle is to the nearest of another type; options are 'pixelwise' (value is calculated for each organelle pixel and then averaged) or 'instance' (value is calculated for each individual organelle and then averaged).  Defaults to 'instance'.

*MDR_type* - Whether the MDR (mead distribution radius) should be calculated using the distance to the centroid or edge of the nucleus; options are 'centroid' or 'nucleus'.  The 'nucleus' option requires nuclei to be measured in the ImageJ analysis step; 'centroid' will fall back to cell centroid if there is no nucleus data, but may be less accurate.  Defaults to 'centroid'.

*exclude_ER_instance* - Exclude metrics related to ER instance segmentation - e.g. average/sd of size, cicrularity, etc.  Metrics looking at the ER overall (area, MDR, CoA, etc.) will still me included.

*exclude_nuclei_metrics* - Exclude all metrics related to nuclei and organelle-nucleus contacts (with the exception of '_nucleus_MDR' for other organelles and contacts).

```{r user_params}
up <- list()#leave this line as-is - this variable will be used to retrieve the values below
up['batch'] <- 1
up['experiment'] <- 2
up['rep'] <- 3
up['cond'] <- 4
up['image'] <- 5


binangle = 5 #degree width of bins for asymmetry coefficient
if((360/binangle) %% 2 != 0){ warning("Binangle must divide 360 into an even number of groups for the following code to work.") }

#metric options
use_normalised_metrics <- TRUE
normalise_organelles_to <- 'cell' #TODO - NOT FULLY IMPLEMENTED
distance_metric <- 'instance'
MDR_type <- 'nucleus' #TODO - manual cell centroid option?
exclude_ER_instance <- TRUE
exclude_nuclei_metrics <- FALSE

```


## ImageJ import and metrics

*R setup*
```{r imagej_setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tcltk)
library(dplyr)
# library(ggplot2)
# library(ggpubr)
# library(patchwork)
# library(corrr)
# library(ggcorrplot)
# library(factoextra)
#TODO - check these are all still used; move PCA-related to vis section

#for advanced metrics
library(terra)

#additional packages specified directly in function calls:
library(rstudioapi)

```

*Select ImageJ Macro output directory*
```{r data_location}
dir <- rstudioapi::selectDirectory(caption = "Select 'anaysis' folder from ImageJ macro")
#TODO - note about running in Rstudio/from other tools - might need to change/specify directory
filesep = .Platform$file.sep
```

*Identify conditions for matching and iterating over ROI/measurement files*
```{r identify_conditions}
files <- list.files(dir, pattern = ('\\.csv$|\\.zip$'), recursive = FALSE, include.dirs = FALSE)
files <- files[-which(grepl('\\_ALL\\_CELL\\_ROIs\\.zip$|\\_ALL\\_ORGANELLE\\_ROIs\\.zip$', files))]
condcells <- unique(gsub('_results.csv||_POOLED_results.csv||_POOLED_ROIs.zip||_ROIs.zip||_summary_results.csv||_slice_labels.csv', '', files))#updated for cell-by-cell processing
```

*Identify and process measurement files for each condition/cell*
```{r process_measurements}
setClass("Measurements", slots = c(data = "data.frame", cells = "data.frame", other = "data.frame"))

#measurements cleaning/categories
process_measurements <- function(dir, condition, file_pattern){
  df <- read.csv(paste0(dir, filesep, condition, file_pattern))
  df$ROI_name <- unlist(regmatches(df$Label, regexec("(?<=:).+(?=:)?", df$Label, perl = T)))
  #TODO - this might break if trailing : missing - dependent on single/multi channel image - should be consistent with macro
  df.cells = df[df$Type == 'cells',]
  df.other = df[df$Type != 'cells',]
  mes <- new("Measurements", data = df, cells = df.cells, other = df.other)
  return(mes)
}

pooled_measurements <- lapply(condcells, function(c){process_measurements(dir, c, "_POOLED_results.csv")})
names(pooled_measurements) <- condcells
individual_measurements <- lapply(condcells, function(c){process_measurements(dir, c, "_results.csv")})
names(individual_measurements) <- condcells
summary_measurements <- lapply(condcells, function(c){process_measurements(dir, c, "_summary_results.csv")})
names(summary_measurements) <- condcells
```

*Identify set of organelles and types of contacts*
```{r organelles_and_contacts}
#Copied from advanced metrics - remove when combining
slicenames.df <- read.csv(paste0(dir, filesep, condcells[1], '_slice_labels.csv'), header = FALSE)#running ONCE - should be consistent across batch
slicenames <- slicenames.df[,2]
organelles <- slicenames.df[slicenames.df$V3!='',2]
contacts <- slicenames.df[slicenames.df$V3=='',2]
contacts <- contacts[which(contacts != 'cell')]
contacts.components <- lapply(contacts, function(c){
  o <- strsplit(c, '')[[1]]
  org <- slicenames.df[slicenames.df$V3 %in% o, 2]
  return(org)
})
names(contacts.components) <- contacts
```


*Combine ImageJ metrics into data frame*
```{r imagej_metrics_data_frame}
metrics <- data.frame(condcell = condcells)

#---define custom variables from file name---
for(param in names(up)){
  metrics[param] <- sapply(condcells, function(c){strsplit(c, "_")[[1]][up[[param]]]})
}
remove(param)
#------------

metrics$cell_area <- sapply(condcells, function(c){ pooled_measurements[[c]]@cells$Area })
metrics$cell_circ <- sapply(condcells, function(c){ pooled_measurements[[c]]@cells$Circ. })

for(org in c(organelles, contacts)){
  metrics[paste0(org, "_area")] <- sapply(condcells, function(c){ other = pooled_measurements[[c]]@other;
                                                          other[other$Type == org,]$Area[1] })#should only have one of each - MIGHT BREAK IF PAIRWISE MISSING - CHECK
  metrics[paste0(org, "_area_norm_cell")] <- metrics[paste0(org, "_area")]/metrics$cell_area
  
  metrics[paste0(org, "_count")] <- sapply(condcells, function(c){ other = individual_measurements[[c]]@other;#FROM INDIVIDUALS DF
                                                          nrow(other[other$Type == org,]) })
  
  metrics[paste0(org, "_size_avg")] <- sapply(condcells, function(c){ other = individual_measurements[[c]]@other;#FROM INDIVIDUALS DF
                                                        mean(other[other$Type == org,]$Area) })
  metrics[paste0(org, "_size_sd")] <- sapply(condcells, function(c){ other = individual_measurements[[c]]@other;#FROM INDIVIDUALS DF
                                                          sd(other[other$Type == org,]$Area) })
  
  metrics[paste0(org, "_circ_avg")] <- sapply(condcells, function(c){ other = individual_measurements[[c]]@other;#FROM INDIVIDUALS DF
                                                        mean(other[other$Type == org,]$Circ.) })
  metrics[paste0(org, "_circ_sd")] <- sapply(condcells, function(c){ other = individual_measurements[[c]]@other;#FROM INDIVIDUALS DF
                                                          sd(other[other$Type == org,]$Circ.) })
}
remove(org)

metrics$total_organelle_area <- sapply(condcells, function(c){df <- summary_measurements[[c]]@data; df[df$Slice == 'organelles','Total.Area'][1]})
metrics$total_organelle_area_norm_cell <- metrics$total_organelle_area/metrics$cell_area
metrics$total_contact_area <- sapply(condcells, function(c){df <- summary_measurements[[c]]@data; df[df$Slice == 'contact_sites','Total.Area'][1]})
metrics$total_contact_area_norm_cell <- metrics$total_contact_area/metrics$cell_area
metrics$total_contact_area_norm_organelles <- metrics$total_contact_area/metrics$total_organelle_area

if('nuclei_area' %in% colnames(metrics)){
  metrics$cytoplasmic_area <- metrics$cell_area - metrics$nuclei_area
  if(normalise_organelles_to %in% c('cytoplasm', 'both')){
    for(org in c(organelles)){
      metrics[paste0(org, "_area_norm_cyto")] <- metrics[paste0(org, "_area")]/metrics$cytoplasmic_area
      #TODO - exclude nuclei at this stage (area related to cytoplasm area)
    }
    remove(org)
    metrics$total_organelle_area_norm_cyto <- metrics$total_organelle_area/metrics$cytoplasmic_area
    metrics$total_contact_area_norm_cyto <- metrics$total_contact_area/metrics$cytoplasmic_area
  }
  
}
```

*Data cleaning*
```{r imagej_metrics_cleaning}
metrics[is.na(metrics)] <- 0

if(normalise_organelles_to %in% c('cytoplasm', 'none')){#remove columns normalised to cell area, except for nuclei
  to_remove <- colnames(metrics)[grepl('_area_norm_cell', colnames(metrics))]
  to_remove <- to_remove[!to_remove %in% nuclei_area_norm_cell]
}

# Additional code here if necessary - e.g. removing bacterial measurements from 'uninfected' cells

```

Optional - passing through additional measurements from ImageJ, with details from file names
remove 'eval=FALSE' if desired
```{r imagej_passthrough, eval = FALSE}
pooled.df <- do.call(rbind, lapply(1:length(pooled_measurements), function(p){ df <- pooled_measurements[[p]]@data; df$filename <- names(pooled_measurements)[p]; return(df) }))
extra_measurements.pooled <- left_join(metrics[, c(names(up), 'condcell')], pooled.df, join_by(condcell == filename))

individual.df <- do.call(rbind, lapply(1:length(individual_measurements), function(i){ df <- individual_measurements[[i]]@data; df$filename <- names(individual_measurements)[i]; return(df) }))
extra_measurements.individual <- left_join(metrics[, c(names(up), 'condcell')], individual.df, join_by(condcell == filename))
remove(pooled.df, individual.df)
```

*Removing unnecessary variables before advanced metrics processing*
```{r imagej_cleanup}
remove(individual_measurements, summary_measurements, slicenames.df)
```

## Advanced metrics

*Constants and functions*
```{r adv_constants_and_functions}
px_round <- function(x,y){
  xr <- round(x, digits = 0)
  yr <- round(y, digits = 0)
  if((x-xr)<0){ xr <- xr - 0.5 }
  else if((x-xr)>=0){ xr <- xr + 0.5 }
  if((y-yr)<0){ yr <- yr - 0.5 }
  else if((y-yr)>=0){ yr <- yr + 0.5 }
  return(c(xr, yr))
}

adv_metrics <- data.frame(
  condcell = condcells
)

norm_values <- data.frame(
  condcell = condcells
)

cent_type <- ifelse('nuclei' %in% slicenames, 'nuclei', 'cells')
#TODO - manual option to calculate from cell centroid?
```

```{r adv_main_loop}

for(c in condcells){
  print(paste0('Processing condition ', c))
  print(Sys.time())
  cell.raster <- rast(paste0(dir, filesep, c, '_stack.tif'))
  names(cell.raster) <- slicenames
  crs(cell.raster) <- 'local' #IMPORTANT - treat as planar coordinates instead of lon/lat
 
  cell.pixels <- as.data.frame(cell.raster, xy = TRUE) %>% filter(cell >0)

  p <- pooled_measurements[[c]]@data
  cell_cent <- p[p$Type == cent_type, c('X', 'Y')]#should be single row - pulling from 'pooled' data frame
  remove(p)

  cell.pixels$dist_to_cent <- abs(sqrt((cell.pixels$x - cell_cent$X[1])^2 + (cell.pixels$y - cell_cent$Y[1])^2))
  cell.pixels$angle_to_cent <- (atan2((cell.pixels$y - cell_cent$Y[1]), (cell.pixels$x - cell_cent$X[1]))*180)/pi
  cell.pixels$anglebin <- cut_width(cell.pixels$angle_to_cent, width = binangle, closed = 'left', boundary = 0)
  
  cell.raster <- extend(cell.raster, c(1,1), fill = 0, NAflag = NA)#adds negative values to left - no xy offset vs cell.pixels; values still match
  #NAflag setting to fix bug where certain cells have all organelle pixels set to NA from 255
  cr <- subst(cell.raster[['cell']], 255, NA)
  
  cell.pixels$dist_to_periphery <- as.data.frame(mask(distance(cr, exclude = 255), cell.raster[['cell']], maskvalues = 0))[,1]#distance to nearest pixel outside cell
  cell.pixels$cent_norm <- cell.pixels$dist_to_cent/(cell.pixels$dist_to_cent + cell.pixels$dist_to_periphery)
  
  if('nuclei' %in% names(cell.raster)){
    cell.pixels$dist_to_nucleus = as.data.frame(mask(distance(cell.raster[['nuclei']], target = 0), cell.raster[['cell']], maskvalues = 0))[,1]#distance to nearest nucleus pixel (0 if overlapping)
    cell.pixels$nuc_norm <- cell.pixels$dist_to_nucleus/(cell.pixels$dist_to_nucleus + cell.pixels$dist_to_periphery)
  }
  
  for(org in c(organelles, contacts)[!c(organelles, contacts) %in% c('cell', 'nuclei')]){#distance maps
    if(sum(cell.pixels[org])>0){
      cell.pixels[paste0("dist_to_", org)] <- as.data.frame(mask(distance(cell.raster[[org]], target = 0), cell.raster[['cell']], maskvalues = 0))[,1]#distance to nearest org pixel (0 if overlapping)
      
      patches <- as.data.frame(patches(cell.raster[[org]], zeroAsNA = TRUE, directions = 8), xy = TRUE)#8 directions (Queen's case) to match ImageJ behaviour
      cell.pixels <- cell.pixels %>% left_join(patches, join_by(x,y))
      cell.pixels[paste0(org, "_patch")] <- cell.pixels$patches #renaming to organelle-specific
      cell.pixels <- cell.pixels %>% select(-patches)
      remove(patches)
    }
    else{
      cell.pixels[paste0("dist_to_", org)] <- NA
      cell.pixels[paste0(org, "_patch")] <- NA
    }
  }
  remove(org)
  
  
  for(o in slicenames){#including cell here - might normalise to cell asymmetry downstream
    df.org <- cell.pixels[cell.pixels[o] > 0,]
    
    if(nrow(df.org) == 0){#if organelle is absent in cell, bypass calculations, assign all metrics to NA - MUST MIRROR ELSE BLOCK TO AVOID MISSING COLUMNS
      adv_metrics[which(adv_metrics$condcell == c), paste0(o, '_centroid_MDR')] <- NA
      if('nuclei' %in% names(cell.raster)){
        adv_metrics[which(adv_metrics$condcell == c), paste0(o, '_nucleus_MDR')] <- NA
      }
      adv_metrics[which(adv_metrics$condcell == c), paste0(o, '_CoA')] <- NA
      if (!o %in% c('nuclei', 'cell')){
        for(o2 in slicenames[!slicenames %in% c('nuclei', 'cell')]){
          adv_metrics[which(adv_metrics$condcell == c), paste0(o, '_pixelwise_mean_dist_to_', o2)] <- NA
          adv_metrics[which(adv_metrics$condcell == c), paste0(o, '_instance_mean_dist_to_', o2)] <- NA
          adv_metrics[which(adv_metrics$condcell == c), paste0(o, '_percent_in_contact_with_', o2)] <- NA
        }
      }
    }
    else{
      #MDR - TODO skip unnecessary calculations based on user parameters  
      adv_metrics[which(adv_metrics$condcell == c), paste0(o, '_centroid_MDR')] <- mean(df.org$cent_norm)
      if('nuclei' %in% names(cell.raster)){
        adv_metrics[which(adv_metrics$condcell == c), paste0(o, '_nucleus_MDR')] <- mean(df.org$nuc_norm)
      }
      
      #coefficient of asymmetry
      tab <- as.data.frame(table(df.org$anglebin))
      tally <- 0
      halfcount <- nrow(tab)/2
      for(n in 1:halfcount){
        a <- abs(tab[n, 'Freq'] - tab[n + halfcount, 'Freq'])
        tally <- tally + a
      }
      CoA <- tally/nrow(df.org)
      adv_metrics[which(adv_metrics$condcell == c), paste0(o, '_CoA')] <- CoA
      
      #distance to nearest
      if (!o %in% c('nuclei', 'cell')){#already calculated distance to nucleus/PM for MDR
        for(o2 in slicenames[!slicenames %in% c('nuclei', 'cell')]){
          adv_metrics[which(adv_metrics$condcell == c), paste0(o, '_pixelwise_mean_dist_to_', o2)] <- mean(pull(df.org, paste0('dist_to_', o2)))
          
          patch_summary <- df.org %>%
            group_by_at(paste0(o, '_patch')) %>%
            summarise_at(paste0('dist_to_', o2), list(min_dist = min))

          adv_metrics[which(adv_metrics$condcell == c), paste0(o, '_instance_mean_dist_to_', o2)] <- mean(patch_summary$min_dist)
          adv_metrics[which(adv_metrics$condcell == c), paste0(o, '_percent_in_contact_with_', o2)] <- (sum(patch_summary$min_dist == 0)/nrow(patch_summary))*100
          #calculated by instance, not area
        }
      }
    }
    
    if (o %in% contacts){
      #pixel counts for normalising contact areas
      o.components <- contacts.components[[o]]
      pix <- nrow(cell.pixels %>% filter(if_any(any_of(o.components), ~.!=0)))
      norm_values[norm_values$condcell == c, o] <- pix
    }
    
  }

  remove(cell.raster, cell.pixels, cell_cent)
  remove(pix, o.components, df.org, o, o2, CoA, tally, a, n, halfcount, tab, cr, patch_summary)
}
remove(c)
print(paste0("Finished ", Sys.time()))
```
*Combining with ImageJ metrics*
```{r full_metrics_join}

full_metrics <- dplyr::left_join(metrics, adv_metrics, by = join_by(condcell == condcell))

```

*Normalising contact areas to areas of involved organelles* TODO - check performance on missing organelles/contacts - should just be 0
```{r adv_contact_norm}
for(c in contacts){
  # full_metrics[paste0(c, '_area_norm')] <- full_metrics[paste0(c, '_area')]/
  full_metrics[paste0(c, '_area_norm')] <- sapply(full_metrics$condcell, function(cc){#TODO - make this cleaner/check order is consistent
    full_metrics[full_metrics$condcell == cc, paste0(c, '_area')]/norm_values[norm_values$condcell == cc, c]
  })
  #remove corresponding cell-/cyto-normalised columns
  full_metrics <- full_metrics[,colnames(full_metrics)[which(!grepl(paste0(c, '_area_norm_'), colnames(full_metrics)))]]
}
remove(c)
```


*Data cleaning*
```{r adv_metrics_cleaning}
#none applicable - NA values at this stage should stay NA
#add relevant code here if e.g. handling infected/uninfected cells differently, flagging 'bystander' cells, etc.
```

*Separating groups of metrics (e.g. raw vs normalised values) and filtering based on user parameters (see top)*
```{r toggle_groups}

metric_names <- colnames(full_metrics)
metric_names <- metric_names[!metric_names %in% names(up)]#excluding user-specified identity metrics

#normalised metrics
metric_names.norm <- metric_names[grepl('_norm', metric_names)]
metric_names.raw <- sub('_norm.+$', '', metric_names.norm)#TODO - sub vs gsub? should only appear once

#pixelwise vs instance wise distance metrics
metric_names.pixelwise <- metric_names[grepl('_pixelwise_mean_dist_to_', metric_names)]
metric_names.instance <- metric_names[grepl('_instance_mean_dist_to_', metric_names)]

#nucleus vs centroid MDR
metric_names.nucMDR <- metric_names[grepl('_nucleus_MDR', metric_names)]
metric_names.centMDR <- metric_names[grepl('_centroid_MDR', metric_names)]



metric_names.unfiltered <- metric_names[!metric_names %in% c(metric_names.norm, 
                                                             metric_names.raw, 
                                                             metric_names.pixelwise, 
                                                             metric_names.instance, 
                                                             metric_names.nucMDR, 
                                                             metric_names.centMDR)]



#Assemble list of columns to keep

metrics_to_keep <- c(names(up),
                     metric_names.unfiltered,
                     if(use_normalised_metrics == TRUE){metric_names.norm} else if(use_normalised_metrics == FALSE){metric_names.raw} else{metric_names.norm},
                     if(distance_metric == 'instance'){metric_names.instance} else if(distance_metric == 'pixelwise'){metric_names.instance} else{metric_names.instance},
                     if(MDR_type == 'nucleus'){metric_names.nucMDR} else if(MDR_type == 'centroid'){metric_names.centMDR} else{metric_names.centMDR}
                     )

full_metrics.filtered <- full_metrics[,metrics_to_keep] #might have issues here if working with very large data sets - consider removing unjoined/unfiltered tables

```
```{r org_exclusions}
#processing after first round of toggle group filtering - might exclude from either/both groups selected

#specific organelle filtering
metric_names.ERinstance <- c('ER_count', 'ER_size_avg', 'ER_size_sd', 'ER_circ_avg', 'ER_circ_sd', metric_names[grepl('^ER_instance', metric_names)])#hard-coded metric names
metric_names.nuclei <- c(metric_names[grepl('^nuclei_', metric_names)], metric_names[grepl('_nuclei$', metric_names)], #TODO - distance to/percent in contact with nuclei metrics not calculated - add?
                         metric_names[grepl(paste0('^', contacts[grepl('n', contacts)], sep = '', collapse = '|'), metric_names)],
                         metric_names[grepl(paste0(contacts[grepl('n', contacts)], '$', sep = '', collapse = '|'), metric_names)])

metrics_to_exclude = c(
                        if(exclude_ER_instance == TRUE){metric_names.ERinstance},
                        if(exclude_nuclei_metrics == TRUE){metric_names.nuclei}
  
                      )

full_metrics.filtered <- full_metrics.filtered[!colnames(full_metrics.filtered) %in% metrics_to_exclude]

```











