---
title: "Organelle Analysis - ImageJ Metrics"
author: "Fiona M. Love"
date: "2025-05-09"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(ggplot2)
library(ggpubr)
library(patchwork)
library(corrr)
library(ggcorrplot)
library(factoextra)

```

#ANALYSIS

*Select ImageJ Macro output directory*
_R VERSION BUG ON WINDOWS 10_ - WON'T OPEN SELECTION WINDOW UNTIL AFTER file.choose() HAS BEEN RUN
https://stackoverflow.com/questions/77984544/choose-dir-function-does-not-work-in-r-v4-3-2
```{r data_location}
dir <- choose.dir() #WINDOWS ONLY - TODO Mac support with tk option - https://stackoverflow.com/questions/48218491/os-independent-way-to-select-directory-interactively-in-r
```

*Identify conditions for matching and iterating over ROI/measurement files*
```{r conditions}
files <- list.files(dir, pattern = ('\\.csv$|\\.zip$'), recursive = FALSE, include.dirs = FALSE)
condcells <- unique(gsub('_results.csv||_POOLED_results.csv||_POOLED_ROIs.zip||_ROIs.zip||_summary_results.csv||_slice_labels.csv', '', files))#updated for cell-by-cell processing
```

*Identify files for each condition/cell*
Can't read in complex ROIs with RImageJROI package - measurements only for now
```{r file_matching}

setClass("Measurements", slots = c(data = "data.frame", cells = "data.frame", other = "data.frame"))

#measurements cleaning/categories
process_measurements <- function(dir, condition, file_pattern){
  df <- read.csv(paste0(dir, '\\', condition, file_pattern))
  df$ROI_name <- unlist(regmatches(df$Label, regexec("(?<=:).+(?=:)?", df$Label, perl = T)))
  #TODO - this might break if trailing : missing - dependent on single/multi channel image - should be consistent with macro
  df.cells = df[df$Type == 'cells',]
  df.other = df[df$Type != 'cells',]
  mes <- new("Measurements", data = df, cells = df.cells, other = df.other)
  return(mes)
}

pooled_measurements <- lapply(condcells, function(c){process_measurements(dir, c, "_POOLED_results.csv")})
names(pooled_measurements) <- condcells
individual_measurements <- lapply(condcells, function(c){process_measurements(dir, c, "_results.csv")})
names(individual_measurements) <- condcells
summary_measurements <- lapply(condcells, function(c){process_measurements(dir, c, "_summary_results.csv")})
names(summary_measurements) <- condcells

```

*Organelle set*
```{r organelles}
organelles <- unique(unlist(lapply(condcells, function(c){ pooled_measurements[[c]]@other$Type })))
pairwise <- list("ng" = c("nuclei", "Golgi"),
"np" = c("nuclei", "peroxisomes"),
"ne" = c("nuclei", "ER"),
"nm" = c("nuclei", "mitochondria"),
"nl" = c("nuclei", "lysosomes"),
"nb" = c("nuclei", "bacteria"),
"gp" = c("Golgi", "peroxisomes"),
"ge" = c("Golgi", "ER"),
"gm" = c("Golgi", "mitochondria"),
"gl" = c("Golgi", "lysosomes"),
"gb" = c("Golgi", "bacteria"),
"pe" = c("peroxisomes", "ER"),
"pm" = c("peroxisomes", "mitochondria"),
"pl" = c("peroxisomes", "lysosomes"),
"pb" = c("peroxisomes", "bacteria"),
"em" = c("ER", "mitochondria"),
"el" = c("ER", "lysosomes"),
"eb" = c("ER", "bacteria"),
"ml" = c("mitochondria", "lysosomes"),
"mb" = c("mitochondria", "bacteria"),
"lb" = c("lysosomes", "bacteria"))
```


*Combine into table of defined measurements*
```{r data_frame}
metrics <- data.frame(condcell = condcells)

#---CUSTOM MATCHING - Tom's data---
#TODO - more efficient
metrics$infection_time <- sapply(condcells, function(c){strsplit(c, "_")[[1]][2]})
metrics$infected <- sapply(condcells, function(c){strsplit(c, "_")[[1]][3]})
metrics$bac_line <- sapply(condcells, function(c){strsplit(c, "_")[[1]][4]})
metrics$cell_line <- sapply(condcells, function(c){strsplit(c, "_")[[1]][5]})
metrics$image <- sapply(condcells, function(c){strsplit(c, "_")[[1]][6]})
metrics$cell_number <- sapply(condcells, function(c){strsplit(c, "_")[[1]][8]})
#------------

metrics$cell_area <- sapply(condcells, function(c){ pooled_measurements[[c]]@cells$Area })
metrics$cell_circ <- sapply(condcells, function(c){ pooled_measurements[[c]]@cells$Circ. })

for(org in c(organelles, names(pairwise))){
  metrics[paste0(org, "_area")] <- sapply(condcells, function(c){ other = pooled_measurements[[c]]@other;
                                                          other[other$Type == org,]$Area[1] })#should only have one of each - MIGHT BREAK IF PAIRWISE MISSING - CHECK
  metrics[paste0(org, "_area_norm_cell")] <- metrics[paste0(org, "_area")]/metrics$cell_area
  metrics[paste0(org, "_count")] <- sapply(condcells, function(c){ other = individual_measurements[[c]]@other;#FROM INDIVIDUALS DF
                                                          nrow(other[other$Type == org,]) })
  if(org %in% names(pairwise)){
    metrics[paste0(org, "_size_avg")] <- sapply(condcells, function(c){ other = individual_measurements[[c]]@other;#FROM INDIVIDUALS DF
                                                          mean(other[other$Type == org,]$Area) })
    metrics[paste0(org, "_size_sd")] <- sapply(condcells, function(c){ other = individual_measurements[[c]]@other;#FROM INDIVIDUALS DF
                                                          sd(other[other$Type == org,]$Area) })
  }
}

metrics$total_organelle_area <- sapply(condcells, function(c){df <- summary_measurements[[c]]@data; df[df$Slice == 'organelles','Total.Area'][1]})
metrics$total_organelle_area_norm_cell <- metrics$total_organelle_area/metrics$cell_area
metrics$total_contact_area <- sapply(condcells, function(c){df <- summary_measurements[[c]]@data; df[df$Slice == 'pairwise_contact_sites','Total.Area'][1]})
metrics$total_contact_area_norm_cell <- metrics$total_contact_area/metrics$cell_area

if('nuclei_area' %in% colnames(metrics)){
  metrics$cytoplasmic_area <- metrics$cell_area - metrics$nuclei_area
}




```



## Data cleaning ##
```{r}
metrics[is.na(metrics)] <- 0
#Tom specific: replaces all bacterial measurements with 0 in uninfected conditions
metrics[which(metrics$infected == 'UNinfected'),] <- metrics[which(metrics$infected == 'UNinfected'),] %>% mutate(across(contains('bacteria'), ~replace(.,1:length(.),0)))
metrics[which(metrics$infected == '1dpi'),]$infected <- 'infected'
```


## Extra Metrics ##
Passing through additional measurements from ImageJ, with details from file names
```{r}
pooled.df <- do.call(rbind, lapply(1:length(pooled_measurements), function(p){ df <- pooled_measurements[[p]]@data; df$filename <- names(pooled_measurements)[p]; return(df) }))
extra_measurements.pooled <- left_join(metrics[, c('infection_time', 'infected', 'bac_line', 'cell_line', 'image', 'cell_number', 'condcell')], pooled.df, join_by(condcell == filename))

individual.df <- do.call(rbind, lapply(1:length(individual_measurements), function(i){ df <- individual_measurements[[i]]@data; df$filename <- names(individual_measurements)[i]; return(df) }))
extra_measurements.individual <- left_join(metrics[, c('infection_time', 'infected', 'bac_line', 'cell_line', 'image', 'cell_number', 'condcell')], individual.df, join_by(condcell == filename))
```



