---
title: "Organelle Analysis - all metrics"
author: "Fiona M. Love"
date: "2025-11-11"
output: html_document
---
# ORGANELLE ANALYSIS

## User parameters

*file name parsing* - each section of the file name delineated by underscores (_) can be passed as a variable representing e.g. timepoint, cell line, experimental condition, etc.  Indicate which place each one is in here, ensuring there are no spaces or other special characters (besides underscores) in the name.
e.g. for something like "HeLa_Ctrl_cell1":
up['cell_line'] <- 1
up['condition'] <- 2
up['cell'] <- 3

*normalise_organelles_to* - Which measurement to use to normalise organelle areas.  Options are 'cell', 'cytoplasm' (cell area - nucleus area), 'both', or 'none'. 'Cytoplasm' or 'both' will only work if nuclei were measured in the ImageJ analysis step, and nucleus will still be normalised to cell area unless 'none' is selected.  Organelle contacts are normalised to the cumulative area of the organelle types involved.

*binangle* - The angle to use to split the cell into slices for calculating coefficient of asymmetry.  Must divide 360 into an even number of groups - e.g. 5, 6, or 10 is fine, but 7 or 8 won't work.

```{r user_params}
up <- list()#leave this line as-is - this variable will be used to retrieve the values below
up['batch'] <- 1
up['experiment'] <- 2
up['rep'] <- 3
up['cond'] <- 4
up['image'] <- 5

normalise_organelles_to <- 'cell'

binangle = 5 #degree width of bins for asymmetry coefficient
if((360/binangle) %% 2 != 0){ warning("Binangle must divide 360 into an even number of groups for the following code to work.") }
```


## ImageJ import and metrics

*R setup*
```{r imagej_setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tcltk)
library(dplyr)
library(ggplot2)
library(ggpubr)
library(patchwork)
library(corrr)
library(ggcorrplot)
library(factoextra)

#additional packages specified directly in function calls:
#rstudioapi
#TODO - check these are all still used; move PCA-related to vis section
```

*Select ImageJ Macro output directory*
```{r data_location}
dir <- rstudioapi::selectDirectory(caption = "Select 'anaysis' folder from ImageJ macro")
#TODO - note about running in Rstudio/from other tools - might need to change/specify directory
filesep = .Platform$file.sep
```

*Identify conditions for matching and iterating over ROI/measurement files*
```{r identify_conditions}
files <- list.files(dir, pattern = ('\\.csv$|\\.zip$'), recursive = FALSE, include.dirs = FALSE)
files <- files[-which(grepl('\\_ALL\\_CELL\\_ROIs\\.zip$|\\_ALL\\_ORGANELLE\\_ROIs\\.zip$', files))]
condcells <- unique(gsub('_results.csv||_POOLED_results.csv||_POOLED_ROIs.zip||_ROIs.zip||_summary_results.csv||_slice_labels.csv', '', files))#updated for cell-by-cell processing
```

*Identify and process measurement files for each condition/cell*
```{r process_measurements}
setClass("Measurements", slots = c(data = "data.frame", cells = "data.frame", other = "data.frame"))

#measurements cleaning/categories
process_measurements <- function(dir, condition, file_pattern){
  df <- read.csv(paste0(dir, filesep, condition, file_pattern))
  df$ROI_name <- unlist(regmatches(df$Label, regexec("(?<=:).+(?=:)?", df$Label, perl = T)))
  #TODO - this might break if trailing : missing - dependent on single/multi channel image - should be consistent with macro
  df.cells = df[df$Type == 'cells',]
  df.other = df[df$Type != 'cells',]
  mes <- new("Measurements", data = df, cells = df.cells, other = df.other)
  return(mes)
}

pooled_measurements <- lapply(condcells, function(c){process_measurements(dir, c, "_POOLED_results.csv")})
names(pooled_measurements) <- condcells
individual_measurements <- lapply(condcells, function(c){process_measurements(dir, c, "_results.csv")})
names(individual_measurements) <- condcells
summary_measurements <- lapply(condcells, function(c){process_measurements(dir, c, "_summary_results.csv")})
names(summary_measurements) <- condcells
```

*Identify set of organelles and types of contacts*
```{r organelles_and_contacts}
#Copied from advanced metrics - remove when combining
slicenames.df <- read.csv(paste0(dir, filesep, condcells[1], '_slice_labels.csv'), header = FALSE)#running ONCE - should be consistent across batch
slicenames <- slicenames.df[,2]
organelles <- slicenames.df[slicenames.df$V3!='',2]
contacts <- slicenames.df[slicenames.df$V3=='',2]
contacts <- contacts[which(contacts != 'cell')]
contacts.components <- lapply(contacts, function(c){
  o <- strsplit(c, '')[[1]]
  org <- slicenames.df[slicenames.df$V3 %in% o, 2]
  return(org)
})
names(contacts.components) <- contacts
```


*Combine ImageJ metrics into data frame*
```{r imagej_metrics_data_frame}
metrics <- data.frame(condcell = condcells)

#---define custom variables from file name---
for(param in names(up)){
  metrics[param] <- sapply(condcells, function(c){strsplit(c, "_")[[1]][up[[param]]]})
}
remove(param)
#------------

metrics$cell_area <- sapply(condcells, function(c){ pooled_measurements[[c]]@cells$Area })
metrics$cell_circ <- sapply(condcells, function(c){ pooled_measurements[[c]]@cells$Circ. })

for(org in c(organelles, contacts)){
  metrics[paste0(org, "_area")] <- sapply(condcells, function(c){ other = pooled_measurements[[c]]@other;
                                                          other[other$Type == org,]$Area[1] })#should only have one of each - MIGHT BREAK IF PAIRWISE MISSING - CHECK
  metrics[paste0(org, "_area_norm_cell")] <- metrics[paste0(org, "_area")]/metrics$cell_area
  
  metrics[paste0(org, "_count")] <- sapply(condcells, function(c){ other = individual_measurements[[c]]@other;#FROM INDIVIDUALS DF
                                                          nrow(other[other$Type == org,]) })
  
  metrics[paste0(org, "_size_avg")] <- sapply(condcells, function(c){ other = individual_measurements[[c]]@other;#FROM INDIVIDUALS DF
                                                        mean(other[other$Type == org,]$Area) })
  metrics[paste0(org, "_size_sd")] <- sapply(condcells, function(c){ other = individual_measurements[[c]]@other;#FROM INDIVIDUALS DF
                                                          sd(other[other$Type == org,]$Area) })
  
  metrics[paste0(org, "_circ_avg")] <- sapply(condcells, function(c){ other = individual_measurements[[c]]@other;#FROM INDIVIDUALS DF
                                                        mean(other[other$Type == org,]$Circ.) })
  metrics[paste0(org, "_circ_sd")] <- sapply(condcells, function(c){ other = individual_measurements[[c]]@other;#FROM INDIVIDUALS DF
                                                          sd(other[other$Type == org,]$Circ.) })
}
remove(org)

metrics$total_organelle_area <- sapply(condcells, function(c){df <- summary_measurements[[c]]@data; df[df$Slice == 'organelles','Total.Area'][1]})
metrics$total_organelle_area_norm_cell <- metrics$total_organelle_area/metrics$cell_area
metrics$total_contact_area <- sapply(condcells, function(c){df <- summary_measurements[[c]]@data; df[df$Slice == 'contact_sites','Total.Area'][1]})
metrics$total_contact_area_norm_cell <- metrics$total_contact_area/metrics$cell_area
metrics$total_contact_area_norm_organelles <- metrics$total_contact_area/metrics$total_organelle_area
#TODO ASAP - NORMALISE TO INVOLVED ORGANELLES FOR EACH CONTACT

if('nuclei_area' %in% colnames(metrics)){
  metrics$cytoplasmic_area <- metrics$cell_area - metrics$nuclei_area
  for(org in c(organelles)){
    metrics[paste0(org, "_area_norm_cyto")] <- metrics[paste0(org, "_area")]/metrics$cytoplasmic_area
    #TODO - conditional on user params
    #TODO - norm nuclei to cell area instead of cyto
  }
  remove(org)
}
```

*Data cleaning*
```{r imagej_metrics_cleaning}
metrics[is.na(metrics)] <- 0
# Additional code here if necessary - e.g. removing bacterial measurements from 'uninfected' cells
```

Optional - passing through additional measurements from ImageJ, with details from file names
remove 'eval=FALSE' if desired
```{r imagej_passthrough, eval = FALSE}
pooled.df <- do.call(rbind, lapply(1:length(pooled_measurements), function(p){ df <- pooled_measurements[[p]]@data; df$filename <- names(pooled_measurements)[p]; return(df) }))
extra_measurements.pooled <- left_join(metrics[, c(names(up), 'condcell')], pooled.df, join_by(condcell == filename))

individual.df <- do.call(rbind, lapply(1:length(individual_measurements), function(i){ df <- individual_measurements[[i]]@data; df$filename <- names(individual_measurements)[i]; return(df) }))
extra_measurements.individual <- left_join(metrics[, c(names(up), 'condcell')], individual.df, join_by(condcell == filename))
remove(pooled.df, individual.df)
```

*Removing unnecessary variables before advanced metrics processing*
```{r imagej_cleanup}
remove(individual_measurements, summary_measurements, slicenames.df)
```

## Advanced metrics
```{r adv_setup, include=FALSE}
library(terra)
```

*Constants and functions*
```{r adv_constants_and_functions}
px_round <- function(x,y){
  xr <- round(x, digits = 0)
  yr <- round(y, digits = 0)
  if((x-xr)<0){ xr <- xr - 0.5 }
  else if((x-xr)>=0){ xr <- xr + 0.5 }
  if((y-yr)<0){ yr <- yr - 0.5 }
  else if((y-yr)>=0){ yr <- yr + 0.5 }
  return(c(xr, yr))
}

adv_metrics <- data.frame(
  condcell = condcells
)

norm_values <- data.frame(
  condcell = condcells
)

cent_type <- ifelse('nuclei' %in% slicenames, 'nuclei', 'cells')
#TODO - manual option to calculate from cell centroid?
```

```{r adv_main_loop}

for(c in condcells){
  print(paste0('Processing condition ', c))
  print(Sys.time())
  cell.raster <- rast(paste0(dir, filesep, c, '_stack.tif'))
  names(cell.raster) <- slicenames
  crs(cell.raster) <- 'local' #IMPORTANT - treat as planar coordinates instead of lon/lat

  cell.pixels <- as.data.frame(cell.raster, xy = TRUE) %>% filter(cell >0)
  
  p <- pooled_measurements[[c]]@data
  cell_cent <- p[p$Type == cent_type, c('X', 'Y')]
  remove(p)

  cell.pixels$dist_to_cent <- abs(sqrt((cell.pixels$x - cell_cent$X[1])^2 + (cell.pixels$y - cell_cent$Y[1])^2))
  cell.pixels$angle_to_cent <- (atan2((cell.pixels$y - cell_cent$Y[1]), (cell.pixels$x - cell_cent$X[1]))*180)/pi
  cell.pixels$anglebin <- cut_width(cell.pixels$angle_to_cent, width = binangle, closed = 'left', boundary = 0)

  
  cell.raster <- extend(cell.raster, c(1,1), fill = 0)#adds negative values to left - no xy offset vs cell.pixels; values still match
  cr <- subst(cell.raster[['cell']], 255, NA)
  
  cell.pixels$dist_to_periphery <- as.data.frame(mask(distance(cr, exclude = 255), cell.raster[['cell']], maskvalues = 0))[,1]#distance to nearest pixel outside cell
  cell.pixels$cent_norm <- cell.pixels$dist_to_cent/(cell.pixels$dist_to_cent + cell.pixels$dist_to_periphery)
  
  if('nuclei' %in% names(cell.raster)){
    cell.pixels$dist_to_nucleus = as.data.frame(mask(distance(cell.raster[['nuclei']], target = 0), cell.raster[['cell']], maskvalues = 0))[,1]#distance to nearest nucleus pixel (0 if overlapping)
    cell.pixels$nuc_norm <- cell.pixels$dist_to_nucleus/(cell.pixels$dist_to_nucleus + cell.pixels$dist_to_periphery)
  }
  
    
  for(org in c(organelles, contacts)[!c(organelles, contacts) %in% c('cell', 'nuclei')]){#distance maps
    if(sum(cell.pixels[org])>0){
      cell.pixels[paste0("dist_to_", org)] <- as.data.frame(mask(distance(cell.raster[[org]], target = 0), cell.raster[['cell']], maskvalues = 0))[,1]#distance to nearest org pixel (0 if overlapping)
      
      patches <- as.data.frame(patches(cell.raster[[org]], zeroAsNA = TRUE ), xy = TRUE)
      cell.pixels <- cell.pixels %>% left_join(patches, join_by(x,y))
      cell.pixels[paste0(org, "_patch")] <- cell.pixels$patches #renaming to organelle-specific
      cell.pixels <- cell.pixels %>% select(-patches)
    }
    else{
      cell.pixels[paste0("dist_to_", org)] <- NA
      cell.pixels[paste0(org, "_patch")] <- NA
    }
  }
  remove(org)
  
  
  # for(o in slicenames[slicenames != 'cell']){
  #   adv_metrics[which(adv_metrics$condcell == c), paste0(o, '_MDR')] <- mean(cell.pixels[cell.pixels[,o]>0,]$norm)
  # }
  for(o in slicenames){#including cell here - might normalise to cell asymmetry downstream
    df.org <- cell.pixels[cell.pixels[o] > 0,]
    
    if(nrow(df.org) == 0){#if organelle is absent in cell, bypass calculations, assign all metrics to NA - MUST MIRROR ELSE BLOCK TO AVOID MISSING COLUMNS
      adv_metrics[which(adv_metrics$condcell == c), paste0(o, '_centroid_MDR')] <- NA
      if('nuclei' %in% names(cell.raster)){
        adv_metrics[which(adv_metrics$condcell == c), paste0(o, '_nucleus_MDR')] <- NA
      }
      adv_metrics[which(adv_metrics$condcell == c), paste0(o, '_CoA')] <- NA
      if (!o %in% c('nuclei', 'cell')){
        for(o2 in slicenames[!slicenames %in% c('nuclei', 'cell')]){
          adv_metrics[which(adv_metrics$condcell == c), paste0(o, '_pixelwise_mean_dist_to_', o2)] <- NA
          adv_metrics[which(adv_metrics$condcell == c), paste0(o, '_instance_mean_dist_to_', o2)] <- NA
          adv_metrics[which(adv_metrics$condcell == c), paste0(o, '_percent_in_contact_with_', o2)] <- NA
        }
      }
    }
    else{
      #MDR  
      adv_metrics[which(adv_metrics$condcell == c), paste0(o, '_centroid_MDR')] <- mean(df.org$cent_norm)
      if('nuclei' %in% names(cell.raster)){
        adv_metrics[which(adv_metrics$condcell == c), paste0(o, '_nucleus_MDR')] <- mean(df.org$nuc_norm)
      }
      
      #coefficient of asymmetry
      tab <- as.data.frame(table(df.org$anglebin))
      tally <- 0
      halfcount <- nrow(tab)/2
      for(n in 1:halfcount){
        a <- abs(tab[n, 'Freq'] - tab[n + halfcount, 'Freq'])
        tally <- tally + a
      }
      CoA <- tally/nrow(df.org)
      adv_metrics[which(adv_metrics$condcell == c), paste0(o, '_CoA')] <- CoA
      
      #distance to nearest
      if (!o %in% c('nuclei', 'cell')){#already calculated distance to nucleus/PM for MDR
        for(o2 in slicenames[!slicenames %in% c('nuclei', 'cell')]){
          adv_metrics[which(adv_metrics$condcell == c), paste0(o, '_pixelwise_mean_dist_to_', o2)] <- mean(pull(df.org, paste0('dist_to_', o2)))
          
          patch_summary <- df.org %>%
            group_by_at(paste0(o, '_patch')) %>%
            summarise_at(paste0('dist_to_', o2), list(min_dist = min))
          
          adv_metrics[which(adv_metrics$condcell == c), paste0(o, '_instance_mean_dist_to_', o2)] <- mean(patch_summary$min_dist)
          adv_metrics[which(adv_metrics$condcell == c), paste0(o, '_percent_in_contact_with_', o2)] <- (sum(patch_summary$min_dist == 0)/nrow(patch_summary))*100
          #calculated by instance, not area
        }
      }
    }
    
    if (o %in% contacts){
      #pixel counts for normalising contact areas
      o.components <- contacts.components[[o]]
      pix <- nrow(cell.pixels %>% filter(if_any(any_of(o.components), ~.!=0)))
      norm_values[norm_values$condcell == c, o] <- pix
    }
    
  }
  
  remove(cell.raster, cell.pixels, cell_cent)
  remove(pix, o.components, df.org, o, CoA, tally, a, n, halfcount, tab, cr, patch_summary)
}
remove(c)
```
*Combining with ImageJ metrics*
```{r full_metrics_join}
full_metrics <- dplyr::left_join(metrics, adv_metrics, by = join_by(condcell == condcell))
```

*Data cleaning*
```{r adv_metrics_cleaning}
#handle NA values - FOR PCA/UMAP/etc. ONLY
full_metrics.forDR <- full_metrics #watch memory usage - fine for now

#MDR - set NA to 1/at periphery
full_metrics.forDR[,grepl('_MDR', colnames(full_metrics.forDR))][is.na(full_metrics.forDR[,grepl('_MDR', colnames(full_metrics.forDR))])] <- 1
#CoA - set NA to 0/perfectly symmetrical
full_metrics.forDR[,grepl('_CoA', colnames(full_metrics.forDR))][is.na(full_metrics.forDR[,grepl('_CoA', colnames(full_metrics.forDR))])] <- 0

#distance_to - set NA to ????? (TODO - REMOVING FOR NOW)
full_metrics.forDR <- full_metrics.forDR[,!grepl('_dist_to_', colnames(full_metrics.forDR))]

#percent in contact with - set NA to 0
full_metrics.forDR[,grepl('_percent_in_contact_with_', colnames(full_metrics.forDR))][is.na(full_metrics.forDR[,grepl('_percent_in_contact_with_', colnames(full_metrics.forDR))])] <- 0

#removing columns with zero variance - technically optional but required for scaling
full_metrics.forDR <- full_metrics.forDR[,which(apply(full_metrics.forDR, 2, var)!=0)]#REMOVES IDENTITY COLUMNS
full_metrics.forDR <- bind_cols(full_metrics[,1:(length(up) + 1)], full_metrics.forDR)#re-adding from full_metrics - CHECK

```

*Normalising contact areas to areas of involved organelles* TODO - check performance on missing organelles/contacts - should just be 0
```{r adv_contact_norm}
for(c in contacts){
  # full_metrics[paste0(c, '_area_norm')] <- full_metrics[paste0(c, '_area')]/
  full_metrics[paste0(c, '_area_norm')] <- sapply(full_metrics$condcell, function(cc){#TODO - make this cleaner/check order is consistent
    full_metrics[full_metrics$condcell == cc, paste0(c, '_area')]/norm_values[norm_values$condcell == cc, c]
  })
}
remove(c)
```

## Visualisation
_TODO_
```{r}
exclude_nuclei_metrics <- TRUE #not excluding nuclear area, CoA, etc.
if(exclude_nuclei_metrics){
  organelles <- organelles[organelles != 'nuclei']
  contacts <- contacts[!grepl('n', contacts)]
}
```

*MEAN DISTANCES HEATMAP*
```{r pull_distance_metrics}
dist_metrics <- full_metrics %>% select(names(up) | contains('_mean_dist_to_'))
dist_metrics.org_to_org <- dist_metrics %>% select(names(up) | starts_with(organelles) & ends_with(organelles))
dist_metrics.contact_to_contact <- dist_metrics %>% select(names(up) | starts_with(contacts) & ends_with(contacts))
dist_metrics.contact_to_org <- dist_metrics %>% select(names(up) | starts_with(contacts) & ends_with(organelles))
dist.metrics.org_to_contacts <- dist_metrics %>% select(names(up) | starts_with(organelles) & ends_with(contacts))
```
```{r reshape_for_heatmap}
#organelle to organelle
org_to_org <- do.call(rbind, lapply(1:nrow(dist_metrics.org_to_org), function(i){
  df <- do.call(rbind, lapply(colnames(dist_metrics.org_to_org)[!colnames(dist_metrics.org_to_org) %in% names(up)], function(col){
          colsplit <- strsplit(col, '_mean_dist_to_')
          from <- colsplit[[1]][1]
          to <- colsplit[[1]][2]
          value <- dist_metrics[i, col]
          d <- data.frame(from = from, to = to, value = value)
          return(d)
        }))
  df[names(up)] <- dist_metrics.org_to_org[rep_len(i, nrow(df)), names(up)]
  return(df)
}))
```
```{r ggplot2_tile_heatmap}
ggplot(org_to_org, aes(y = from, x = to, fill = value)) + geom_tile() + facet_wrap(~cond)
```
